Verifying Fine-Grained and Lock-Free Binary Search Trees

### 2022-01-19 concurrency/Verification of Concurrent Binary Search Tree.tex
clightgen'd up to CompCert 3.10
also patched for 64-bit

\subsubsection{Insert}
An outline of the C code for the insert function is shown in figure \ref{insert-proof}. When a thread inserts a key $k$ into a tree, it first crawls the tree to find the right position for $k$ using the hand-over-hand locking mechanism, acquiring the lock for the next node before releasing the lock for the current node. If it encounters a node with key $k$, then it simply changes the value at that node. Otherwise, it will eventually reach a leaf node, where it creates two new empty leaf nodes, inserts the key and value into the current node, and links the newly created leaf nodes to the current node. The atomic specification for insert method is $$\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle$$
The \texttt{insert} function should atomically add the key $k$ with value $v$ to the tree at $p$.

The key to the verification of the C code is the loop invariant for the top-level loop. We define the loop invariant for the insert function as follows: 
\begin{align*} &\mathsf{insert\_inv}(b, x, g, g\_root) \triangleq\ \exists\ \mathit{lock},\mathit{g\_current},\mathit{np},\mathit{range},\mathit{info},\ (x\in \mathit{range})\ \land \ \ghost{\mathsf{my\_half}(\mathit{range},\mathit{info})}{g\_current}\ *\ R\ \mathit{np}\\& * \mathsf{lock\_inv}(\mathit{lock},\mathit{lsh2},R')\ *\ \nodeboxrep(b,g,\mathit{g\_root})\ *\ \mathsf{atomic\_shift} (P_p,E_i,E_o,Q_p,Q) \end{align*}   
where $P_p$ and $Q_p$ are $\treerep(t,g)$ and $\treerep(t[k\mapsto v],g)$ respectively. The resources $R$, parameterized by a node pointer $\mathit{np}$, describes the concrete information about a node. Also, $x$ is the key to be inserted that must be in the bound $\mathsf{range}$.  This loop invariant has some existential variables which characterize the state of each iteration of the loop.

Figure \ref{insertproof} shows the $\texttt{insert}$ function with separation logic annotations. The proof starts with $\nodeboxrep$ and $\mathsf{atomic\_shift}$ as the precondition. After acquiring the lock for the root node, the thread accesses the information inside the lock invariant of the root node. In order to prove $\mathsf{while}$ loop, we show that the precondition satisfies the $\mathsf{insert\_inv}$, then prove that the loop body preserves the loop invariant (cases inside $\mathsf{else if}$ clauses at line 15 and 17 in \ref{insert}).   Once we locate the empty node for inserting the new key-value pair (inside the first $\mathsf{if}$ clause), we open the global invariant encoded in the $\mathsf{atomic\_shift}$, create the ghost names $\texttt{g1}$ and $\texttt{g2}$ for two new empty child nodes of current node, and use the $\mathsf{sync\_commit}$ rule---we have reached the linearization point and completed the insertion. The premise of $\mathsf{sync\_commit}$ requires that we have enough information to know that inserting the key at this point will change the abstract state from some unknown $t$ to $t[k\mapsto v]$, which we show in the second part of the proof.

The other branch of the code-level proof is the case where the key to be inserted already exists in the tree (the last $\texttt{else}$ clause in the code). %The modification of the global state of the tree t is shown in \ref{extract_insert2}. In that case, the old value $v'$ associated with key is overridden by new value $v$ keeping a key $40$ and child pointers as as they are. This is also the \emph{linearization point} of the $\mathsf{insert}$ operation. We augment the same $\emph{extract}$ lemma with the clause that states the scenario in \ref{extract_insert2}, and prove the lemma in the same induction over the tree $t$.
In this case, we change the value at the node but leave the rest of the tree intact, and this is the linearization point for the insertion. In the second part of the proof, we will show that this case also transforms $t$ to $t[k\mapsto v]$.

In the other two cases, we compare the key
of the current node with the new key \texttt{x} and move to the left or right child accordingly. Here we again need to access the atomic shift, to ensure that the range of the node we move to is included in the range of the current node. We can use $\mathsf{sync\_rollback}$ to retrieve this information without making any changes to the abstract state of the tree. %For the left branch case, we accomplish the proof by the following lemma:
%\begin{verbatim}
%Lemma in_tree_left_range:
%  forall (B: Type) (b: tree -> B -> mpred) (Q : B → mpred) (x x0: Z) (g g_root : gname)
%    (inv_names : invG) (v: val) (g_in ga gb: gname) (r a: node_info),
%    check_key_exist' x r.1 = true -> r.2 = Some (Some (x0, v, ga, gb)) -> x < x0 ->
%    atomic_shift (λ BST : tree, !! sorted_tree BST && tree_rep2 g g_root BST) ∅ ⊤
%                 b Q * my_half g_in gsh1 r * in_tree g g_in * my_half ga gsh1 a
%    |-- atomic_shift (λ BST : tree, !! sorted_tree BST && tree_rep2 g g_root BST) ∅ ⊤
%    b Q * my_half g_in gsh1 r *
%    (EX ba, !! (less_than_equal ba r.1.1 = true /\
%                range_inclusion a.1 (ba, Finite_Integer x0) = true) &&
%            (in_tree g g_in * my_half ga gsh1 (ba, Finite_Integer x0, a.2))).  
%\end{verbatim}
%The lemma for the right branch case is similar. Both use the
%$\texttt{sync\_rollback}$ lemma described in section \ref{atomicity}.

\begin{figure}[htb]
\centering
\includegraphics[width=150mm,scale=0.5]{FIG/extract_insert.png}
\caption{A visual depiction of the change in global state of BST during $\mathsf{insert(k,v)}$ operation }
\label{extract_insert}
\end{figure}
\begin{figure}[htb]
\centering
\includegraphics[width=150mm,scale=0.5]{FIG/extract_insert_2.png}
\caption{A visual depiction of the change in global state of BST during $\mathsf{insert(k,v)}$ operation, where key $k$ already exists in the tree $t$ }
\label{extract_insert2}
\end{figure}
\begin{figure}[htp]
\begin{subfigure}[t]{1\textwidth}
 $$\left\{\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
 \vspace*{-20pt}
\begin{lstlisting}[language = C,numbers = none]
insert(treebox t, int x, void *value)
  tree_t *tgt = *t;
  acquire(tgt->lock)
 \end{lstlisting}  
 $$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}((-\infty,+\infty),info)}{g\_root}\ *\ R\ b\ *\ \mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\
 \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} insert\_inv \end{array}\right\}$$ 
  \begin{lstlisting}[language = C,numbers = none]
  for(;;) {
       \end{lstlisting}   
   $$\left\{\begin{array}{l} insert\_inv \end{array}\right\} \triangleq \left\{\begin{array}{l}(x\in range)\land \ghost{\mathsf{my\_half}(range,info)}{g\_current}*\ R\ np\ \\*\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\}$$
      \begin{lstlisting}[language = C,numbers = none]
    p = tgt->t;
    if (p==NULL)
      tree_t *p1, *p2 = malloc();
      p = malloc(); tgt->t = p;
      p->key=x; p->value=value; p->left=p1; p->right=p2;
           \end{lstlisting} 
  $$\left\{\begin{array}{l} \ghost{\mathsf{own}\ (a1)}{g1}\ *\ \ghost{\mathsf{own}\ (a2)}{g2}\ *\ \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ \\*\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit}}$$
$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
 \vspace*{-10pt}
        \begin{lstlisting}[language = C,numbers = none]
      release2(l);
         \end{lstlisting}
       $$\left\{\begin{array}{l} \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
        \vspace*{-10pt}
         \begin{lstlisting}[language = C,numbers = none]
      return;
     else 
        ....
       else 
      	p->value=value;
      	\end{lstlisting} 
$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *
\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \\\nodeboxrep(b,g\_root)\ *\ \mathsf{atomic\_shift}(P_p,E_i,E_o,Q_p,Q)\end{array}\right\} \Rrightarrow{\textbf{sync\_commit}}$$
$$\left\{\begin{array}{l} \ghost{\mathsf{my\_half}(range,None)}{g\_current}*\ R\ np\ *\mathsf{lock\_inv}(lock,lsh2,R')\ *\ \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
         \vspace*{-10pt}
      	\begin{lstlisting}[language = C, numbers = none]
        release2(l);
        \end{lstlisting}
        $$\left\{\begin{array}{l}  \nodeboxrep(b,g\_root)\ *\ Q\end{array}\right\}$$
         \vspace*{-10pt}
        \begin{lstlisting}[language = C, numbers = none] 
      	return;
\end{lstlisting}
\end{subfigure}
\caption{The $\texttt{insert}$ function annotated with separation logic specification}
\label{insertproof}
\end{figure}

\paragraph{Changing the Global State} describe ``extract lemmas'' here

>>>>

\subsubsection{Insert}
An outline of the C code for the insert function is shown in figure \ref{bst-conc}. When a thread inserts a key $k$ into a tree, it first crawls the tree to find the right position for $k$ using hand-over-hand locking. If it encounters a node with key $k$, then it simply changes the value at that node. Otherwise, it will eventually reach a leaf node, where it creates two new empty leaf nodes, inserts the key and value into the current node, and links the newly created leaf nodes to the current node. The atomic specification for insert method is $$\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle$$
The \lstinline{insert} function should atomically add the key-value pair $(k, v)$ to the map implemented by the BST at $p$.

The main loop of the \lstinline{insert} function traverses the tree to find the node where $k$ should be inserted. At the beginning of each iteration, we are at some node $n$ and hold the lock for that node, giving us the lock variant $\mathsf{node\_inv}((l, r), c)$ where $(l, r)$ is the range of keys in the subtree rooted at $n$. The loop invariant holds that $k \in (l, r)$, that is, the key $k$ always belongs in the subtree of the current node. This guarantees us that if we reach an empty leaf, that leaf is exactly the right place to add a node with key $k$.

The function has two possible linearization points: after line 23, where we finish adding a node $(k, v)$ to the tree, and after line 41, where we change the value of an existing node with key $k$ to $v$. In both cases, we use the \textsf{sync\_commit} rule to atomically update the abstract state of the data structure from $\treerep m$ to $\treerep (\mathrm{insert}\ m\ k\ v)$, by showing that the update to the tree implementation corresponds to a local change in the abstract tree that implements the mathematical $\mathsf{insert}$. In the first case, the local change is the insertion of $(k, v)$ at an empty leaf whose range includes $k$, and the addition of two new empty leaves as its children; in the second case, we simply change the value of an internal node whose key is already $k$. Our higher-level proofs, then, need to show that these two changes are both valid operations on ghost trees that yield the desired insertion in the global state.

\begin{figure}[htb]
\centering
\includegraphics[width=150mm,scale=0.5]{FIG/extract_insert.png}
\caption{A visual depiction of the change in global state of BST during $\mathsf{insert(k,v)}$ operation }
\label{extract_insert}
\end{figure}
\begin{figure}[htb]
\centering
\includegraphics[width=150mm,scale=0.5]{FIG/extract_insert_2.png}
\caption{A visual depiction of the change in global state of BST during $\mathsf{insert(k,v)}$ operation, where key $k$ already exists in the tree $t$ }
\label{extract_insert2}
\end{figure}

\paragraph{Changing the Global State} describe ``extract lemmas'' here---how much detail?
